\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-03-20 Paper/paper.tex
change in representation yet again

\texttt{GIFT}\cite{gift:2017}, first presented in the \textit{CHES 2017}
cryptographic hardware and embedded systems conference, is a lightweight block
cipher based on a previous design called \texttt|PRESENT|, developed in 2007. Its
goal is to offer maximum security while being extremely light on resources.
Modern battery-powered devices like RFID tags or low-latency operations like
on-the-fly disc encryption present strong hardware and power constraints. GIFT
aims to be a simple, low-energy cipher suited for these kinds of applications.

@@

This approach requires $47$ operations, meaning all four permutations require
over $150$ operations which would present a major bottleneck to the round
function. We can improve on this by working on multiple message blocks at once
and using the aforementioned \texttt{vperm} instruction to implement the bit
shuffling. We then need only four \texttt{TBL} instructions for the complete
diffusion layer.

@@

\[
    \begin{array}{c|llllllll}
        n & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 \\
        \hline
        V_0 & A56 & A48 & A40 & A32 & A24 & A16 & A8 & A0 \\
        V_1 & A57 & A49 & A41 & A33 & A25 & A17 & A9 & A1 \\
        V_2 & A58 & A50 & A42 & A34 & A26 & A18 & A10 & A2 \\
        V_3 & A59 & A51 & A43 & A35 & A27 & A19 & A11 & A3 \\
        V_4 & A60 & A52 & A44 & A36 & A28 & A20 & A12 & A4 \\
        V_5 & A61 & A53 & A45 & A37 & A29 & A21 & A13 & A5 \\
        V_6 & A62 & A54 & A46 & A38 & A30 & A22 & A14 & A6 \\
        V_7 & A63 & A55 & A47 & A39 & A31 & A23 & A15 & A7
    \end{array}
\]

\[
    \begin{array}{c|llllllll}
        n & 15 & 14 & 13 & 12 & 11 & 10 & 9 & 8 \\
        \hline
        V_0 & B56 & B48 & B40 & B32 & B24 & B16 & B8 & B0 \\
        V_1 & B57 & B49 & B41 & B33 & B25 & B17 & B9 & B1 \\
        V_2 & B58 & B50 & B42 & B34 & B26 & B18 & B10 & B2 \\
        V_3 & B59 & B51 & B43 & B35 & B27 & B19 & B11 & B3 \\
        V_4 & B60 & B52 & B44 & B36 & B28 & B20 & B12 & B4 \\
        V_5 & B61 & B53 & B45 & B37 & B29 & B21 & B13 & B5 \\
        V_6 & B62 & B54 & B46 & B38 & B30 & B22 & B14 & B6 \\
        V_7 & B63 & B55 & B47 & B39 & B31 & B23 & B15 & B7
    \end{array}
\]

@@

One thing to take note of is the original permutation values only showing where
a given byte should land, not which byte belongs to a certain position - i.e.
for $P_0$, byte 1 should land in position 12, but the byte belonging to
position 1 is byte 4. Because \texttt{vperm} works in the latter way, we have
to do some trivial rearrangements. In addition, because data for each slice is
split between two registers $(V_0,V_4),(V_1,V_5),\dots$, the \texttt{vperm}
operation needs to use these two registers as a data source. Using two data
sources is supported by the \texttt{TBL} instruction by means of an additional
parameter.

Assuming the correct permutation values are put into registers
$V_9,V_{10},\dots,V_{16}$, this now allows us to compute the permutation layer for
all 16 blocks in only eight permutation instructions plus four additional vector
copy instructions used for saving the original values of $V_0,V_1,V_2,V_3$
before the permutation is applied.

\begin{alignat*}{2}
    &\texttt{MOV V17, V0}\qquad &&\texttt{MOV V18, V1} \\
    &\texttt{MOV V19, V2}\qquad &&\texttt{MOV V20, V3} \\
    &\texttt{TBL V0, \{V0, V4\}, V9}\qquad &&\texttt{TBL V1, \{V1, V5\}, V10} \\
    &\texttt{TBL V2, \{V2, V6\}, V9}\qquad &&\texttt{TBL V3, \{V3, V7\}, V10} \\
    &\texttt{TBL V4, \{V17, V4\}, V9}\qquad &&\texttt{TBL V5, \{V18, V5\}, V10} \\
    &\texttt{TBL V6, \{V19, V6\}, V9}\qquad &&\texttt{TBL V7, \{V20, V7\}, V10}
\end{alignat*}

>>>>

\texttt{GIFT}\cite{gift:2017}, first presented in the \textit{CHES 2017}
cryptographic hardware and embedded systems conference, is a lightweight block
cipher based on a previous design called \texttt{PRESENT}, developed in 2007. Its
goal is to offer maximum security while being extremely light on resources.
Modern battery-powered devices like RFID tags or low-latency operations like
on-the-fly disc encryption present strong hardware and power constraints. GIFT
aims to be a simple, low-energy cipher suited for these kinds of applications.

@@

This approach requires $47$ operations, meaning all four permutations require
over $150$ operations which would present a major bottleneck to the round
function. We can improve on this by working on multiple message blocks at once
and using the aforementioned \texttt{vperm} instruction to implement the bit
shuffling. We then need only four instructions for the complete diffusion
layer.

@@

\[
    \scriptsize
    \begin{array}{c|llllllll|llllllll}
        n & 15 & 14 & 13 & 12 & 11 & 10 & 9 & 8 & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 \\
        \hline
        V_0 & B56 & B48 & B40 & B32 & B24 & B16 & B8 & B0 & A56 & A48 & A40 & A32 & A24 & A16 & A8 & A0 \\
        V_1 & B57 & B49 & B41 & B33 & B25 & B17 & B9 & B1 & A57 & A49 & A41 & A33 & A25 & A17 & A9 & A1 \\
        V_2 & B58 & B50 & B42 & B34 & B26 & B18 & B10 & B2 & A58 & A50 & A42 & A34 & A26 & A18 & A10 & A2 \\
        V_3 & B59 & B51 & B43 & B35 & B27 & B19 & B11 & B3 & A59 & A51 & A43 & A35 & A27 & A19 & A11 & A3 \\
        V_4 & B60 & B52 & B44 & B36 & B28 & B20 & B12 & B4 & A60 & A52 & A44 & A36 & A28 & A20 & A12 & A4 \\
        V_5 & B61 & B53 & B45 & B37 & B29 & B21 & B13 & B5 & A61 & A53 & A45 & A37 & A29 & A21 & A13 & A5 \\
        V_6 & B62 & B54 & B46 & B38 & B30 & B22 & B14 & B6 & A62 & A54 & A46 & A38 & A30 & A22 & A14 & A6 \\
        V_7 & B63 & B55 & B47 & B39 & B31 & B23 & B15 & B7  & A63 & A55 & A47 & A39 & A31 & A23 & A15 & A7
    \end{array}
\]

Although this would already work, we prefer to have only bits of the same
messages in each register - otherwise the permutation would need to operate on
two source registers with the added requirement of storing the pre-permutation
values for the first four registers, slowing down the round function through
superfluous load/stores. This transformation is trivial by use of
\texttt{TBL} with two data source operands.

@@

Another thing to take note of is the original permutation values only showing
where a given byte should land, not which byte belongs to a certain position -
i.e. for $P_0$, byte 1 should land in position 12, but the byte belonging to
position 1 is byte 4. Because \texttt{TBL} works in the latter way, we have
to do some trivial rearrangements.

Assuming the correct permutation values are put into registers
$V_8,V_{10},\dots,V_{15}$, this now allows us to compute the permutation layer
for all 16 blocks in only eight permutation instructions.

\begin{alignat*}{2}
    &\texttt{TBL V0, V0, V8}\qquad &&\texttt{TBL V1, V1, V9} \\
    &\texttt{TBL V2, V2, V10}\qquad &&\texttt{TBL V3, V3, V11} \\
    &\texttt{TBL V4, V4, V12}\qquad &&\texttt{TBL V5, V5, V13} \\
    &\texttt{TBL V6, V6, V14}\qquad &&\texttt{TBL V7, V7, V15}
\end{alignat*}
